---
title: "Cause of Death- {gganimate} with a Bar Graph"
description: |
  This is part 1 of final project in [Communicating and Transforming Data](https://uo-datasci-specialization.github.io/c2-communicate_transform_data/). 
author:
  - name: Teresa Chen
    url: https://teresashchen.netlify.com/
date: 03-08-2019
output:
  distill::distill_article:
    self_contained: false
---

 
The dataset, called “Leading Causes of Death: United States” was published by [Centers for Disease Control and Prevention](https://www.data.gov/). The data presents age-adjusted death rates for 10 leading causes of death in the United States from 1999 to 2016. According to the [publisher](https://catalog.data.gov/dataset/age-adjusted-death-rates-for-the-top-10-leading-causes-of-death-united-states-2013), the data are based on the information from all resident death certificates filed in the 50 states and the District of Columbia using demographic and medical characteristics.


The R project aims to answer the following three questions by demonstrating three main data vidualizations:  

Q1. How do leading causes of death change over past `r 2016-1999+1` years?  
Q2. What are changing patterns of each leading cause of death over years?  
Q3. What are distinct causes of death in each state of the United States?  


This post presents the first part of the project, in which data visualization for Q1 is shown along with steps leading to the final plot.  

First, required packages are as follows:  

```{r required-package, echo=TRUE, eval=FALSE}

install.packages("rio")
install.packages("here")
install.packages("tidyverse")
install.packages("gganimate")
install.packages("paletteer")
install.packages("maps")
install.packages("transformr")
install.packages("skimr")
install.packages("janitor")

```

Second, load packages and set up the global settings:  

```{r setup, echo=TRUE}

library(tidyverse)
library(rio)
library(here)
library(gganimate)
library(paletteer)
library(maps)
library(skimr)
library(janitor)

knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```


## Data preparation 

Import data with clean variable names and name it as `df`. 

```{r importdata}

df <- import(here::here("data", "Leading_Causes_of_Death__United_States.csv"),
              setclass = "tbl_df") %>% 
  janitor::clean_names()

str(df)
```

## Data overview

The original dataset contains 6 variables, including `r colnames(df)`. Before visiualizing data, I always find useful to clean dataset and do the following steps:  

1. `dplyr::select` useful variables  
  
2. `dplyr::mutate` variable types accordingly, including using `forcat::as.factor`, and
change value names or transform data value if necessary.  
  
3. `dplyr::filter` to keep useful observation

```{r tidydf}


tidy_df <- df %>% 
  dplyr::select(year, cause_name, state, deaths, age_adjusted_death_rate) %>%   
  mutate(cause = as_factor(cause_name),
         state = as_factor(state),
         
         #transform rate to decimal value not %
         rate = age_adjusted_death_rate/100) %>%   
  dplyr::select(year, state, cause, deaths, rate) %>% 
  
  # filter out "all causes"
  filter(cause != "All causes",
         state != "United States") %>% 
  
  # replace "CLRD" with its whole name
  mutate(cause = replace(as.character(cause), cause == "CLRD", 
                         "Chronic lower respiratory diseases"))

skim(tidy_df)
```

## Summary of data

1. No missing data.  
2. There are `r length(unique(tidy_df$cause))` causes of death, including 
`r unique(tidy_df$cause)`.
3. There are `r length(unique(tidy_df$state))` states. 
4. Dataframe has data from 1999 to 2016 (`r 2016-1999+1` years).

## First plot

The first plot is intended to deliver a message to **general population** and to answer *how leading death causes change over years*.  

```{r plot1_v1, fig.width=15, fig.height=20, layout = "l-body-outset"}

plot1_v1 <- tidy_df %>% 
  
  # calculate yearly total deaths from all states
  group_by(year, cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  
  # plot
  ggplot(aes(x = fct_reorder(cause, deaths_by_year_cause), 
             y = deaths_by_year_cause/1000000)) +
    geom_col(fill = "steelblue",
             alpha = 0.7) +
    scale_y_continuous(expand = c(0, 0)) +
    coord_flip() +
    facet_wrap(~year, ncol = 3) +
    theme_minimal(base_size = 20) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold")) +
    labs(y = "No. of deaths (in millions)",
         x = "Causes of deaths",
         title = "Heart disease is a leading cause of death over years",
         caption = "Source: Centers for Disease Control and Prevention")
         

plot1_v1
```


```{r plot1_v1_r1, fig.width=15, fig.height=20, layout = "l-body-outset"}

tidy_df %>% 
  
  # calculate yearly total deaths from all states
  group_by(year, cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  
  # plot
  ggplot(aes(x = fct_reorder(cause, deaths_by_year_cause), 
             y = deaths_by_year_cause/1000000,
             fill = cause)) +
    geom_col(alpha = 0.7) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_brewer(palette = "BrBG") +
    coord_flip() +
    facet_wrap(~year, ncol = 3) +
    theme_minimal(base_size = 20) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"),
          legend.position = "none") +
    labs(y = "No. of deaths (in millions)",
         x = "Causes of deaths",
         title = "Heart disease is a leading cause of death over years",
         caption = "Source: Centers for Disease Control and Prevention")
         

```

