---
title: "How do causes of death change from 1999 to 2016?"
description: |
  Cause of Death- {gganimate} with a Bar Graph 
author:
  - name: Teresa Chen
    url: https://teresashchen.github.io/blog/
date: 03-08-2019
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3    
categories:
  - dataviz
  - rstats
---

This is part 1 of the final project in [Communicating and Transforming Data](https://uo-datasci-specialization.github.io/c2-communicate_transform_data/). In addition to answering the questions regarding causes of death using real data, I also hope to facilitate data analysis using R by documenting the steps leading towards the final results. 

The dataset, called “Leading Causes of Death: United States” was published by [Centers for Disease Control and Prevention](https://www.data.gov/). The data presents age-adjusted death rates for 10 leading causes of death in the United States from 1999 to 2016. According to the [publisher](https://catalog.data.gov/dataset/age-adjusted-death-rates-for-the-top-10-leading-causes-of-death-united-states-2013), the data are based on the information from all resident death certificates filed in the 50 states and the District of Columbia using demographic and medical characteristics.

The R project aims to answer the following three questions by demonstrating three main data vidualizations:  

Q1. How do leading causes of death change over the past `r 2016-1999+1` years?  
Q2. What are the changing patterns of each leading cause of death over the years?  
Q3. What are the distinct causes of death in each state in the United States?  

This post presents the first part of the project, in which data visualization for Q1 is shown along with the steps leading to the final plot.  

First, required packages are as follows:  

```{r required-package, echo=TRUE, eval=FALSE}

install.packages("rio")
install.packages("here")
install.packages("tidyverse")
install.packages("skimr") 
install.packages("janitor")
install.packages("gganimate")
install.packages("paletteer") # for colors

```

Second, load packages and set up the global settings:  

```{r setup, echo=TRUE}

library(tidyverse)
library(rio)
library(here)
library(skimr)
library(janitor)
library(gganimate)
library(paletteer)

knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```


## Data preparation 

Import data with clean variable names and name it as `df`. 

```{r importdata}

df <- import(here::here("data", "Leading_Causes_of_Death__United_States.csv"),
              setclass = "tbl_df") %>% 
  janitor::clean_names()

str(df)
```

## Data overview

The original dataset contains 6 variables, including `r colnames(df)`. Before visiualizing data, I always find it useful to clean the dataset and do the following steps:  

1. `dplyr::select` useful variables  
  
2. `dplyr::mutate` variable types accordingly, including using `forcat::as.factor`, and
change value names or transform data value if necessary.  
  
3. `dplyr::filter` to keep useful observation

```{r tidydf}


tidy_df <- df %>% 
  dplyr::select(year, cause_name, state, deaths, age_adjusted_death_rate) %>%   
  mutate(cause = as_factor(cause_name),
         state = as_factor(state),
         #transform rate to decimal value not %
         rate = age_adjusted_death_rate/100) %>% 
  dplyr::select(year, state, cause, deaths, rate) %>%
  # filter out "all causes"
  filter(cause != "All causes",
         state != "United States") %>%
  # replace "CLRD" with its whole name
  mutate(cause = replace(as.character(cause), cause == "CLRD", 
                         "Chronic lower respiratory diseases"))

skim(tidy_df)
```

## Summary of data

1. No missing data.  
2. There are `r length(unique(tidy_df$cause))` causes of death, including 
`r unique(tidy_df$cause)`.
3. There are `r length(unique(tidy_df$state))` states. 
4. Dataframe has data from 1999 to 2016 (`r 2016-1999+1` years).


## Data visulization  

The first #dataviz of this R project is intended to deliver a message to the **general population** and to answer *how leading causes of death change over the years*. In addition to showing the final plot, I include all iterations of plots and broke down the problem into sub-problems/versions leading up to my final plot.  


### First plot- bar chart 

```{r plot1_v1, fig.width=15, fig.height=20, layout = "l-body-outset"}

plot1_v1 <- tidy_df %>%
  # calculate yearly total deaths from all states
  group_by(year, cause) %>%
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  # plot
  ggplot(aes(x = fct_reorder(cause, deaths_by_year_cause), 
             y = deaths_by_year_cause/100000)) +
    geom_col(fill = "steelblue",
             alpha = 0.7) +
    scale_y_continuous(expand = c(0, 0)) +
    coord_flip() +
    facet_wrap(~year, ncol = 3) +
    theme_minimal(base_size = 20) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold")) +
    labs(y = "No. of deaths (in 100k)",
         x = "Causes of death",
         title = "Heart Disease is a Leading Cause of Death Over Time",
         caption = "Source: Centers for Disease Control and Prevention")
         

plot1_v1
```

### Second plot- bar chart with colors  

From the graph it is easy to tell heart disease is the leading cause of death over the years while cancer is a close second.

A drawback of this chart is the use of a single color. 

My original plan was to use `colorblindr::scale_fill_OkabeIto`, but it has only 8 colors while there are 10 causes of death to be plotted. I end up using one of the diverging palette from [colorbrewer 2.0](http://colorbrewer2.org/#type=diverging&scheme=BrBG&n=10).


```{r plot1_v1_r1, fig.width=15, fig.height=20, layout = "l-body-outset"}

  
tidy_df %>%
  # calculate yearly total deaths from all states
  group_by(year, cause) %>%
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  # plot
  ggplot(aes(x = fct_reorder(cause, deaths_by_year_cause), 
             y = deaths_by_year_cause/100000,
             fill = cause)) +
    geom_col(alpha = 0.7) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_brewer(palette = "BrBG") +
    coord_flip() +
    facet_wrap(~year, ncol = 3) +
    theme_minimal(base_size = 20) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"),
          legend.position = "none") +
    labs(y = "No. of deaths (in 100k)",
         x = " ",
         title = "Heart Disease is a Leading Cause of Death Over Time",
         caption = "Source: Centers for Disease Control and Prevention")
         

```


### Third plot- **animated** bar chart

Although the top 2 causes of death have a clear trend and ranking over the years, a total of 18 facets make it hard to compare the amounts of death and rankings over the years. Therefore I used `{gganimate}` package to condense the information to a single year and animate changes over the years. 


```{r plot1_v2, layout = "l-body-outset"}

plot1_v2 <- tidy_df %>% 
  group_by(year, cause) %>%
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  # plot
  ggplot(aes(x = fct_reorder(cause, deaths_by_year_cause), 
             y = deaths_by_year_cause/100000,
             fill = cause)) +
    geom_col(alpha = 0.7) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_brewer(palette = "BrBG") +
    coord_flip() +
    theme_minimal(base_size = 20) +
    theme(panel.grid.minor = element_blank(),
          plot.margin = margin(1, 0, 1, 0.3, "cm"),
          legend.position = "none") +
    transition_time(year) +
    labs(title = "Top 10 Causes of Death",
         subtitle = "Year: {round(frame_time)}",
         caption = "Source: Centers for Disease Control and Prevention",
         y = "No. of deaths (in 100k)",
         x = "")


animate(plot1_v2, duration = 25, nframes = 100, end_pause = 10,
        width = 800, height = 600, renderer = gifski_renderer("ani_bar.gif"))

```


### Forth plot- **ranking** chart

Although the animation shows clear changes in the amount of deaths across the years, the rankings did not demonstrat properly on the y-axis even using `fct_reorder`. Therefore, here is another version:


```{r plot1_v3, layout = "l-body-outset"}

plot1_v3 <- tidy_df %>% 
  group_by(year, cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>% 
  # creat rank per year
  group_by(year) %>% 
  arrange(desc(deaths_by_year_cause)) %>% 
  mutate(rank = row_number()) %>% 
  arrange(year) %>% 
  ungroup() %>% 
  #plot
  ggplot(aes(x = '{frame_time}', y = rank)) +
  geom_text(aes(label = cause, 
                color = cause,
                group = cause),
            size = 8) +
  scale_x_discrete(labels = NULL) +
  scale_y_reverse() +
  scale_color_paletteer_d(rcartocolor, Vivid) +
  transition_states(year, 
                    transition_length = 1,
                    state_length = 3) +
  ease_aes('sine-in-out') +
  theme_void(base_size = 28) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Top 10 Causes of Death in {closest_state}",
       caption = "Source: Centers for Disease Control and Prevention") +
  guides(color = "none")
  
  
animate(plot1_v3, duration = 30, nframes = 300,
        width = 800, height = 600, end_pause = 10, 
        renderer = gifski_renderer("ani_rank.gif"))
```


### Fifth chart- animated **sorted** bar chart  

This version of data visualization allows us to see change in rankings over the years. For example, Alzheimer's disease increased from 2003-2004 and from 2006-2007. The previous version, however, provides more info in terms of how far ahead one cause is from another cause. 

This [stackoverflow thread](https://stackoverflow.com/questions/53162821/animated-sorted-bar-chart-with-bars-overtaking-each-other) taught me how I could change the ranks of causes of death over y-axis.  

Here is the static version with only the year 2016 shown in the figure:

```{r plot1_v4_preview, fig.height=10, fig.width= 12, echo=FALSE, layout = "l-body-outset", preview=TRUE}

tidy_df %>% 
  filter(year == 2016) %>% 
  group_by(cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  # create rank variable and variable label
  mutate(rank = row_number(-deaths_by_year_cause),
         val_lbl = paste0(" ", deaths_by_year_cause)) %>% 
  ggplot(aes(x = rank, group = cause, fill = cause, color = cause)) +
    geom_tile(aes(y = deaths_by_year_cause/2,
                  height = deaths_by_year_cause,
                  width = 0.9),
              alpha = 0.8) +
    coord_flip(clip = "off", expand = FALSE) +
    scale_x_reverse() +
    scale_fill_brewer(palette = "BrBG") +
    scale_color_brewer(palette = "BrBG") +
    geom_text(aes(y = 0, label = paste(cause, " "), vjust = 0.2, hjust = 1), 
              color = "black", size = 5) +
    geom_text(aes(y = deaths_by_year_cause, label = val_lbl), hjust = 0,
              color = "black") +
    guides(color = FALSE, fill = FALSE) +
    theme_classic(base_size = 25) +
    theme(plot.title = element_text(hjust = 0, vjust = 5, size = 22),
          plot.caption = element_text(size = 13), 
          axis.ticks.y = element_blank(),
          axis.text.y = element_blank(),
          axis.text.x = element_text(size = 15),
          plot.margin = margin(1, 2, 1, 8, "cm")) +
    labs(title='Top 10 Causes of Death in 2016', 
         x = "", 
         y = "",
         caption = "Source: Centers for Disease Control and Prevention")

```

Let's see how I make it alive with `{gganimate}`! 

```{r plot1_v4, layout = "l-body-outset"}

plot1_ani <- tidy_df %>% 
  group_by(year, cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  group_by(year) %>% 
  # create rank variable and variable label
  mutate(rank = row_number(-deaths_by_year_cause),
         val_lbl = paste0(" ", deaths_by_year_cause)) %>% 
  ungroup() %>% 
  ggplot(aes(x = rank, group = cause, fill = cause, color = cause)) +
    geom_tile(aes(y = deaths_by_year_cause/2,
                  height = deaths_by_year_cause,
                  width = 0.9),
              alpha = 0.8) +
    coord_flip(clip = "off", expand = FALSE) +
    scale_x_reverse() +
    scale_fill_brewer(palette = "BrBG") +
    scale_color_brewer(palette = "BrBG") +
    geom_text(aes(y = 0, label = paste(cause, " "), vjust = 0.2, hjust = 1), 
              color = "black", size = 5) +
    geom_text(aes(y = deaths_by_year_cause, label = val_lbl), hjust = 0,
              color = "black") +
    guides(color = FALSE, fill = FALSE) +
    theme_classic() +
    theme(plot.title = element_text(hjust = 0, vjust = 0.5, size = 22),
          plot.caption = element_text(size = 13), 
          axis.ticks.y = element_blank(),
          axis.text.y = element_blank(),
          axis.text.x = element_text(size = 15),
          plot.margin = margin(1, 2, 1, 5.5, "cm")) +
    labs(title='Top 10 Causes of Death in {closest_state}', 
         x = "", 
         y = "",
         caption = "Source: Centers for Disease Control and Prevention") +
    transition_states(year, transition_length = 4, state_length = 1) +
    ease_aes("cubic-in-out")



animate(plot1_ani, 200, fps = 10, duration = 30, end_pause = 10,
        width = 800, height = 600, 
        renderer = gifski_renderer("plot1_ani.gif"))
  

```

The key feature here is to use `geom_tile` to allow the bars to slide positions. Through this plot, I've learned the use of `clip = "off"` in `coord_flip`, which allows drawing of data points anywhere on the plot and the use of `expand = FALSE` to add a small expansion factor to the limits to ensure that data and axes don't overlap.

## Conclusion  

This post is intended to deliver an important message to the general community.

Heart disease is a leading cause of death over the years and cancer is the second highest. The number of deaths from the top two causes are three times more than the sum of deaths from all other causes each year. It seems like Alzheimer's disease is climbing up on its ranking over the years. 


