---
title: "Cause of Death- {gganimate} with a Bar Graph"
description: |
  This is part 1 of final project in [Communicating and Transforming Data](https://uo-datasci-specialization.github.io/c2-communicate_transform_data/). 
author:
  - name: Teresa Chen
    url: https://teresashchen.netlify.com/
date: 03-08-2019
output:
  distill::distill_article:
    self_contained: false
    
categories:
  - dataviz
  - rstats
---

 
The dataset, called “Leading Causes of Death: United States” was published by [Centers for Disease Control and Prevention](https://www.data.gov/). The data presents age-adjusted death rates for 10 leading causes of death in the United States from 1999 to 2016. According to the [publisher](https://catalog.data.gov/dataset/age-adjusted-death-rates-for-the-top-10-leading-causes-of-death-united-states-2013), the data are based on the information from all resident death certificates filed in the 50 states and the District of Columbia using demographic and medical characteristics.


The R project aims to answer the following three questions by demonstrating three main data vidualizations:  

Q1. How do leading causes of death change over past `r 2016-1999+1` years?  
Q2. What are changing patterns of each leading cause of death over years?  
Q3. What are distinct causes of death in each state of the United States?  


This post presents the first part of the project, in which data visualization for Q1 is shown along with steps leading to the final plot.  

First, required packages are as follows:  

```{r required-package, echo=TRUE, eval=FALSE}

install.packages("rio")
install.packages("here")
install.packages("tidyverse")
install.packages("skimr") 
install.packages("janitor")
install.packages("gganimate")
install.packages("paletteer") # for colors

```

Second, load packages and set up the global settings:  

```{r setup, echo=TRUE}

library(tidyverse)
library(rio)
library(here)
library(skimr)
library(janitor)
library(gganimate)
library(paletteer)

knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```


## Data preparation 

Import data with clean variable names and name it as `df`. 

```{r importdata}

df <- import(here::here("data", "Leading_Causes_of_Death__United_States.csv"),
              setclass = "tbl_df") %>% 
  janitor::clean_names()

str(df)
```

## Data overview

The original dataset contains 6 variables, including `r colnames(df)`. Before visiualizing data, I always find useful to clean dataset and do the following steps:  

1. `dplyr::select` useful variables  
  
2. `dplyr::mutate` variable types accordingly, including using `forcat::as.factor`, and
change value names or transform data value if necessary.  
  
3. `dplyr::filter` to keep useful observation

```{r tidydf}


tidy_df <- df %>% 
  dplyr::select(year, cause_name, state, deaths, age_adjusted_death_rate) %>%   
  mutate(cause = as_factor(cause_name),
         state = as_factor(state),
         #transform rate to decimal value not %
         rate = age_adjusted_death_rate/100) %>% 
  dplyr::select(year, state, cause, deaths, rate) %>%
  # filter out "all causes"
  filter(cause != "All causes",
         state != "United States") %>%
  # replace "CLRD" with its whole name
  mutate(cause = replace(as.character(cause), cause == "CLRD", 
                         "Chronic lower respiratory diseases"))

skim(tidy_df)
```

## Summary of data

1. No missing data.  
2. There are `r length(unique(tidy_df$cause))` causes of death, including 
`r unique(tidy_df$cause)`.
3. There are `r length(unique(tidy_df$state))` states. 
4. Dataframe has data from 1999 to 2016 (`r 2016-1999+1` years).


## Data visulization  

The first #dataviz of this R project is intended to deliver a message to **general population** and to answer *how leading death causes change over years*. In addition to showing the final plot, I include all iterations of plots and broke down the problem into sub-problems/versions leading up to my final plot.  


### First plot- bar chart 

```{r plot1_v1, fig.width=15, fig.height=20, layout = "l-body-outset"}

plot1_v1 <- tidy_df %>%
  # calculate yearly total deaths from all states
  group_by(year, cause) %>%
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  # plot
  ggplot(aes(x = fct_reorder(cause, deaths_by_year_cause), 
             y = deaths_by_year_cause/100000)) +
    geom_col(fill = "steelblue",
             alpha = 0.7) +
    scale_y_continuous(expand = c(0, 0)) +
    coord_flip() +
    facet_wrap(~year, ncol = 3) +
    theme_minimal(base_size = 20) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold")) +
    labs(y = "No. of deaths (in 100k)",
         x = "Causes of deaths",
         title = "Heart disease is a leading cause of death over years",
         caption = "Source: Centers for Disease Control and Prevention")
         

plot1_v1
```

### Second plot- bar chart with colors  

From the graph it is easy to tell heart disease is shown to be a leading cause of death over years while cancer is a close number 2.

One feedback from a peer reviewer is the single color on the chart. One palette I originally plan to use is `colorblindr::scale_fill_OkabeIto`, but it has only 8 colors while there are 10 causes of death to be plot. I end up using one of the diverging palette from [colorbrewer 2.0](http://colorbrewer2.org/#type=diverging&scheme=BrBG&n=10).



```{r plot1_v1_r1, fig.width=15, fig.height=20, layout = "l-body-outset"}

  
tidy_df %>%
  # calculate yearly total deaths from all states
  group_by(year, cause) %>%
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  # plot
  ggplot(aes(x = fct_reorder(cause, deaths_by_year_cause), 
             y = deaths_by_year_cause/100000,
             fill = cause)) +
    geom_col(alpha = 0.7) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_brewer(palette = "BrBG") +
    coord_flip() +
    facet_wrap(~year, ncol = 3) +
    theme_minimal(base_size = 20) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"),
          legend.position = "none") +
    labs(y = "No. of deaths (in 100k)",
         x = " ",
         title = "Heart disease is a leading cause of death over years",
         caption = "Source: Centers for Disease Control and Prevention")
         

```


### Third plot- **animated** bar chart

Although the top 2 causes have a clear trend and rank over years, a total of 18 facets make it hard to compare amounts of death and ranks of causes across years. Therefore I used `{gganimate}` package to condense information to a single year and animate changes over years. 


```{r plot1_v2, layout = "l-body-outset"}

plot1_v2 <- tidy_df %>% 
  group_by(year, cause) %>%
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  # plot
  ggplot(aes(x = fct_reorder(cause, deaths_by_year_cause), 
             y = deaths_by_year_cause/100000,
             fill = cause)) +
    geom_col(alpha = 0.7) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_brewer(palette = "BrBG") +
    coord_flip() +
    theme_minimal(base_size = 20) +
    theme(panel.grid.minor = element_blank(),
          plot.margin = margin(1, 0, 1, 0.3, "cm"),
          legend.position = "none") +
    transition_time(year) +
    labs(title = "Top 10 Causes of Deaths",
         subtitle = "Year: {round(frame_time)}",
         caption = "Source: Centers for Disease Control and Prevention",
         y = "No. of deaths (in 100k)",
         x = "")


animate(plot1_v2, duration = 25, nframes = 100, end_pause = 10,
        width = 800, height = 600, renderer = gifski_renderer("ani_bar.gif"))

```


### Forth plot- **rank** chart

Although the animation shows clear changes in amount of death across years, the  rank of causes did not demonstrated properly on y-axis even using `fct_reorder`. Therefore, here is another version:


```{r plot1_v3, layout = "l-body-outset"}

plot1_v3 <- tidy_df %>% 
  group_by(year, cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>% 
  # creat rank per year
  group_by(year) %>% 
  arrange(desc(deaths_by_year_cause)) %>% 
  mutate(rank = row_number()) %>% 
  arrange(year) %>% 
  ungroup() %>% 
  #plot
  ggplot(aes(x = '{frame_time}', y = rank)) +
  geom_text(aes(label = cause, 
                color = cause,
                group = cause),
            size = 8) +
  scale_x_discrete(labels = NULL) +
  scale_y_reverse() +
  scale_color_paletteer_d(rcartocolor, Vivid) +
  transition_states(year, 
                    transition_length = 1,
                    state_length = 3) +
  ease_aes('sine-in-out') +
  theme_void(base_size = 28) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Top 10 Causes of Death in {closest_state}",
       caption = "Source: Centers for Disease Control and Prevention") +
  guides(color = "none")
  
  
animate(plot1_v3, duration = 30, nframes = 300,
        width = 800, height = 600, end_pause = 10, 
        renderer = gifski_renderer("ani_rank.gif"))
```


### Fifth chart- animated **sorted** bar chart  

This version of data visualization allows us to see changes in rank of causes over years, for example, the rank of Alzheimer's disease increased from 2003-2004 and 2006-2007. Nevertheless, My peer reviewr and I both prefer the previous version because it provides more info in terms of how far ahead one cause is from another cause. 

I spent a good amount of time to figure out how I could change the ranks of cause over y-axis. This [stackoverflow thread](https://stackoverflow.com/questions/53162821/animated-sorted-bar-chart-with-bars-overtaking-each-other) exactly teaches me how to do it.  

Here is the static version with only year 2016 shown in the figure:

```{r plot1_v4_preview, fig.height=8, fig.width= 12, echo=FALSE, layout = "l-body-outset", preview=TRUE}

tidy_df %>% 
  filter(year == 2016) %>% 
  group_by(cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  # create rank variable and variable label
  mutate(rank = row_number(-deaths_by_year_cause),
         val_lbl = paste0(" ", deaths_by_year_cause)) %>% 
  ggplot(aes(x = rank, group = cause, fill = cause, color = cause)) +
    geom_tile(aes(y = deaths_by_year_cause/2,
                  height = deaths_by_year_cause,
                  width = 0.9),
              alpha = 0.8) +
    coord_flip(clip = "off", expand = FALSE) +
    scale_x_reverse() +
    scale_fill_brewer(palette = "BrBG") +
    scale_color_brewer(palette = "BrBG") +
    geom_text(aes(y = 0, label = paste(cause, " "), vjust = 0.2, hjust = 1), 
              color = "black", size = 5) +
    geom_text(aes(y = deaths_by_year_cause, label = val_lbl), hjust = 0,
              color = "black") +
    guides(color = FALSE, fill = FALSE) +
    theme_classic() +
    theme(plot.title = element_text(hjust = 0, vjust = 5, size = 22),
          plot.caption = element_text(size = 13), 
          axis.ticks.y = element_blank(),
          axis.text.y = element_blank(),
          axis.text.x = element_text(size = 15),
          plot.margin = margin(1, 2, 1, 8, "cm")) +
    labs(title='Top 10 Causes of Death in 2016', 
         x = "", 
         y = "",
         caption = "Source: Centers for Disease Control and Prevention")

```

Let's see how I make it alive with `{gganimate}`! 

```{r plot1_v4, layout = "l-body-outset"}

plot1_ani <- tidy_df %>% 
  group_by(year, cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>%
  group_by(year) %>% 
  # create rank variable and variable label
  mutate(rank = row_number(-deaths_by_year_cause),
         val_lbl = paste0(" ", deaths_by_year_cause)) %>% 
  ungroup() %>% 
  ggplot(aes(x = rank, group = cause, fill = cause, color = cause)) +
    geom_tile(aes(y = deaths_by_year_cause/2,
                  height = deaths_by_year_cause,
                  width = 0.9),
              alpha = 0.8) +
    coord_flip(clip = "off", expand = FALSE) +
    scale_x_reverse() +
    scale_fill_brewer(palette = "BrBG") +
    scale_color_brewer(palette = "BrBG") +
    geom_text(aes(y = 0, label = paste(cause, " "), vjust = 0.2, hjust = 1), 
              color = "black", size = 5) +
    geom_text(aes(y = deaths_by_year_cause, label = val_lbl), hjust = 0,
              color = "black") +
    guides(color = FALSE, fill = FALSE) +
    theme_classic() +
    theme(plot.title = element_text(hjust = 0, vjust = 0.5, size = 22),
          plot.caption = element_text(size = 13), 
          axis.ticks.y = element_blank(),
          axis.text.y = element_blank(),
          axis.text.x = element_text(size = 15),
          plot.margin = margin(1, 2, 1, 5.5, "cm")) +
    labs(title='Top 10 Causes of Death in {closest_state}', 
         x = "", 
         y = "",
         caption = "Source: Centers for Disease Control and Prevention") +
    transition_states(year, transition_length = 4, state_length = 1) +
    ease_aes("cubic-in-out")



animate(plot1_ani, 200, fps = 10, duration = 30, end_pause = 10,
        width = 800, height = 600, 
        renderer = gifski_renderer("plot1_ani.gif"))
  

```

The key feature here is to use `geom_tile` to allow the bars slide positions. Other things I've learned through this plot are the use of `clip = "off"` in `coord_flip`, which allows drawing of data points anywhere on the plot and the use of `expand = FALSE` to add a small expansion factor to the limits to ensure that data and axes don't overlap.

## Conclusion  

The first #dataviz of this R project is intended to deliver a message to general audience. Heart disease is a leading cause of death over years and cancer is the second of highest. The top two causes individually take three times more of deaths compare to all the other causes every year. It seems Alzheimer's disease increases its rank on the top 10 causes of death over years. 


