---
title: "How does R&D expenditure changes?"
description: |
  Funding resources are not looking bright for biomechanists.
author:
  - name: Teresa Chen
    url: https://teresashchen.github.io/blog/
date: 03-22-2019
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3     
categories:
  - dataviz
  - rstats
  - biomechanics
---

This is one of my participation in [#TidyTuesday](https://github.com/rfordatascience/tidytuesday) data visualization challenge. The data/visualization, nevertheless, is highgly related to every researcher. I am interested to know **how R&D expenditure changes over years**. Since I am a biomechanist working in academia, I also want to know **how the funding situation loos like for certain organizations where biomechanists usually get funded**.

<aside>
#TidyTuesday is a weekly event on twitter, where R users can practice their wrangling and data visualization skills with real-world data.
</aside>


## About the data

The original data come from American Association for the Advancement of Science Historical Trends. The dataset I used for datavisualization is a *cleaned* version, which means I did not have to organize/clean several excel files to get the final dataset. Thanks to R4DS community!


<aside>
Details regarding the data can be found [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-02-12).
</aside>

Let's do some #dataviz, shall we?

Here are packages I used for this data visulization:

```{r setup, include=TRUE, echo=TRUE}
library(tidyverse)
library(rio)
library(paletteer)
library(gghighlight)
library(cowplot)

knitr::opts_chunk$set(echo = TRUE)
```

## Data preparation

### Import the data

```{r importdata}

fed_rd <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-12/fed_r_d_spending.csv")

```

### Tidy the data

Since I am not familiar with abbreviations of all funding departments, I decide to add their names to the dataset. 

I also devided R&D expenditure to GDP to get funding percentage relative to the yearly economy.

```{r tidydata}

abbname <- tibble(abb = c("DOD - Department of Defense",
                      "NASA - National Aeronautics and Space Administration",
                      "DOE - Department of Energy",
                      "HHS - Department of Health and Human Services",
                      "NIH - National Institute of Health",
                      "NSF - National Science Foundation",
                      "USDA - US Department of Agriculture",
                      "Interior - Department of Interior",
                      "DOT - Deparment of Transportation",
                      "EPA - Environmental Protection Agency",
                      "DOC - Department of Corrections",
                      "DHS - Department of Homeland Security",
                      "VA - Department of Veterands Affairs",
                      "Other - other research and development spending")) %>% 
  separate(abb, into = c("abb", "name"), sep = " - ")


feddf <- left_join(fed_rd, abbname, by = c("department" = "abb"))

feddf <- feddf %>% 
  mutate(percent = rd_budget/gdp)

# skimr::skim(fed_rd)

```

## Data visualization 1- heat map

```{r plot1, fig.width = 25, fig.height = 10, layout = "l-page", preview=TRUE}


#plot heatmap

feddf %>% 
  ggplot(aes(x = year, y = fct_reorder(name, percent))) +
  geom_tile(aes(fill = percent*100),
            color = "white",
            alpha = 0.8) +
  scale_fill_paletteer_c(scico, roma) +
  coord_equal() +
  scale_x_continuous(breaks = seq(1976, 2017, 5),
                     expand = c(0, 0)) +
  theme_minimal(base_size = 25) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5,
                                  vjust = 10,
                                  face = "bold")) +
  labs(fill = "",
       x = "",
       y = "",
       title = "R&D funding as a percent of GDP",
       caption = "American Association for the Advancement of Science")

ggsave("plot0.png", width = 25, height = 10, dpi = 150, units = "in")  
```

DoD stands out from all the other agencies to fund more R&D projects, although the percentage is still low relative to GDP.

## Data visualization 2

### line graph 1

Here I focused on graphing percentage R&D expenditure and used `{gghightlight}` to stand out four departments where biomechanists typically get grant from. 

```{r plot2, fig.height = 15, fig.width = 15}

# plot line graph with percentage

plot1 <- feddf %>% 
  ggplot(aes(x = year, y = percent, color = name)) +
  geom_line(size = 2) +
  scale_color_paletteer_d(ochRe, nolan_ned) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = seq(1976, 2018, 5),
                     expand = c(0, 0)) +
  gghighlight(name == "National Institute of Health" |
              name == "National Science Foundation" |
              name == "Department of Health and Human Services" |
              name == "Department of Defense",
              use_group_by = FALSE,
              use_direct_label = FALSE) +
  theme_minimal(base_size = 18) +
  labs(y = "% of GDP",
       x = "Year",
       color = "Department") +
  theme(legend.justification = c(1, 0.1))

plot1

```

### line graph 2

Instead of using percentage, I plotted R&D expenditure in dollars.

```{r fig.height = 15, fig.width = 15}

# plot line graph with dollars

plot2 <- feddf %>% 
  ggplot(aes(x = year, y = rd_budget/1000000000, color = name)) +
  geom_line(size = 2) +
  scale_color_paletteer_d(ochRe, nolan_ned) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(breaks = seq(1976, 2018, 5),
                     expand = c(0, 0)) +
  gghighlight(name == "National Institute of Health" |
              name == "National Science Foundation" |
              name == "Department of Health and Human Services" |
              name == "Department of Defense",
              use_group_by = FALSE,
              use_direct_label = FALSE) +
  theme_minimal(base_size = 18) +
  labs(y = "$USD (billions)",
       x = "",
       color = "Department") +
  theme(legend.justification = c(1, 0.1))

plot2
```

### Cow plot

It's simply a combination plot of line plot 1 and line plot 2. I think putting 2 plots together can make the story more complete.  Highly recommend `{cowplot}` here to make combining plots super straightforward!

```{r fig.height = 15, fig.width = 15}

# comebine 2 line graphs

plot3 <- plot_grid(plot2, plot1, labels = c("A", "B"), nrow = 2, align = "v")
title <- ggdraw() + 
  draw_label("R&D expenditure \nHighlight 4 agencies where biomechanists usually get grant from",
             fontface = 'bold',
             size = 20)

plot3 <- plot_grid(title, plot3, ncol = 1, rel_heights = c(0.1, 1))

save_plot("plot3.png", plot3,
          ncol = 2, # we're saving a grid plot of 2 columns
          nrow = 2, # and 2 rows
          # each individual subplot should have an aspect ratio of 1.3
          base_aspect_ratio = 1.3
          )

plot3
```

Although the expenditure (money-wise) went up, the percentage relative to GDP went down. 
