---
title: "Cause of Death- {gghighlight} with a Line Graph"
description: |
  This is part 2 of final project in [Communicating and Transforming Data](https://uo-datasci-specialization.github.io/c2-communicate_transform_data/). 
author:
  - name: Teresa Chen
    url: https://teresashchen.netlify.com/
date: 03-12-2019
output:
  distill::distill_article:
    self_contained: false
    
categories:
  - dataviz
  - rstats
---

The R project aims to answer the following three questions by demonstrating three main data vidualizations:  

Q1. How do leading causes of death change over past `r 2016-1999+1` years?  
Q2. What are changing patterns of each leading cause of death over years?  
Q3. What are distinct causes of death in each state of the United States?  

This post presents the second part of the project, in which data visualization for Q2 is shown. For steps leading to the plot, including data preparation and tidy data, please refer to [part 1](https://teresashchen.github.io/blog/posts/2019-03-08-cause-of-death-gganimate-with-a-bar-graph/) of the post.  

Here are required packages:

```{r required-package, echo=TRUE, eval=FALSE}

install.packages("rio")
install.packages("here")
install.packages("tidyverse")
install.packages("paletteer")
install.packages("janitor")
install.packages("gghighlight")
install.packages("colorblinr")

```

Here is a global settings:

```{r setup, echo=TRUE}

library(tidyverse)
library(rio)
library(here)
library(paletteer)
library(janitor)
library(gghighlight)
library(colorblindr)

knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```


```{r import_tidy_data, include=FALSE}

df <- import(here::here("data", "Leading_Causes_of_Death__United_States.csv"),
              setclass = "tbl_df") %>% 
  janitor::clean_names()

tidy_df <- df %>% 
  dplyr::select(year, cause_name, state, deaths, age_adjusted_death_rate) %>%   
  mutate(cause = as_factor(cause_name),
         state = as_factor(state),
         #transform rate to decimal value not %
         rate = age_adjusted_death_rate/100) %>% 
  dplyr::select(year, state, cause, deaths, rate) %>%
  # filter out "all causes"
  filter(cause != "All causes",
         state != "United States") %>%
  # replace "CLRD" with its whole name
  mutate(cause = replace(as.character(cause), cause == "CLRD", 
                         "Chronic lower respiratory diseases"))
```

## Data visualization 

The second plot is for **policy-maker or health-related research** and is a quick summary plot regarding changes in causes of death over years. 

### First plot- line graph with annotation

```{r plot2,fig.height=10, fig.width=12, layout = "l-body-outset"}

# create data for labeling
df_plot2 <-tidy_df %>% 
  group_by(year, cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths))

  
tidy_df %>% 
  group_by(year, cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>%   
  #plot
  ggplot(aes(x = year, y = deaths_by_year_cause/10000, color = cause)) +
    geom_line(size = 1) +
    scale_x_continuous(breaks = seq(2000, 2016, by = 2),
                       expand = c(0, 0)) +
    scale_y_log10(expand = c(0, 0), breaks = c(3, 5, 10, 20, 30, 60, 100),
                  limits = c(2, 100)) + 
    scale_color_paletteer_d(rcartocolor, Vivid) +
    theme_classic(base_size = 15) +
    geom_text(data = filter(df_plot2, year == 2016),aes(label = cause),
              nudge_x = 5.5, hjust = 1, size = 4) +
    guides(color = "none") +
    theme(panel.grid.minor = element_blank(),
          legend.key.size = unit(3, 'lines'),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(face = "italic")) +
    labs(title = "Causes of Death Across Years",
         subtitle = "Heart disease and cancer are top 2 causes of deaths.\nAlzheimer's disease increases rapidly.",
         x = "Year",
         y = "No. of deaths (in 10k)",
         caption = "Source: Centers for Disease Control and Prevention ")

              
  
  
```

From the plot it is easy to tell the changes in number of death across years. Some dieases show an increased number (suicide, kidney disease, diabetes, Alzheimer's disease, chronic lower respiratory disease, unintentional injuries), some disease show a decreased number (influenza and pneumonia, stroke, heart disease) while some does not change to much (cancer).

The cons of this plot is the texts annotating causes are overlapped. I have tried to maximize the figure height and space out each line to avoid the issue without distorting a figure ratio to musch, but the problem remains the same to some texts. For example, "Influenza and pneumonia" is on top of "Kidney disease". So how about getting rid of the text then? 


### Second plot- line plot with a color legend

```{r plot2_v2, fig.width=18, fig.height=13, layout = "l-body-outset"}
tidy_df %>% 
  group_by(year, cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>% 
  mutate(cause = factor(cause, 
                        levels = c("Heart disease", "Cancer", "Unintentional injuries",
                                   "Chronic lower respiratory diseases", "Stroke", "Alzheimer's disease", "Diabetes", "Influenza and pneumonia", "Kidney disease" , "Suicide"  
                                   ))) %>% 
  ggplot(aes(x = year, y = deaths_by_year_cause/10000, color = cause)) +
    geom_line(size = 2) +
    scale_x_continuous(breaks = seq(2000, 2016, by = 2), expand = c(0, 0)) +
    scale_y_log10(expand = c(0, 0), breaks = c(3, 5, 10, 20, 30, 60, 100),
                  limits = c(2, 100)) + 
    scale_color_paletteer_d(rcartocolor, Vivid) +
    theme_classic(base_size = 26) +
    theme(panel.grid.minor = element_blank(),
          legend.key.size = unit(3, 'lines'),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(face = "italic")) +
    labs(title = "Causes of Death Across Years",
         subtitle = "Heart disease and cancer are top 2 causes of deaths.\nAlzheimer's disease increases rapidly.",
         x = "Year",
         y = "No. of deaths (in 10k)",
         caption = "Source: Centers for Disease Control and Prevention ",
         color = "")
```


Although the texts of each cause are legible in the color legend now, a total of 10 colors definitly increases audiences' cognitive loads. The other concern is that there is no color-blind-friendly qualitative palette out there that has more than 10 colors. Therefore, inspired by peer reviewrs' suggestion, I decide to use `{gghighlight}` to select three causes of interest and plot these three causes with color-blind-friendly palette `colorblindr::scale_color_OkabeIto()`in the final version of plot below. Note that `scale_color_OkabeIto` needs additional two packages `{cowplot}` and `{colorspace}` to make the palette work. Here is the [detail](https://www.rdocumentation.org/packages/colorblindr/versions/0.1.0). 


```{r plot2_v3, fig.width=10, fig.height=8, layout = "l-body-outset", preview=TRUE}

tidy_df %>% 
  group_by(year, cause) %>% 
  summarise(deaths_by_year_cause = sum(deaths)) %>% 
  mutate(cause = factor(cause, 
                        levels = c("Heart disease", "Cancer", "Unintentional injuries",
                                   "Chronic lower respiratory diseases", "Stroke", "Alzheimer's disease", "Diabetes", "Influenza and pneumonia", "Kidney disease" , "Suicide"  
                                   ))) %>%  
  ggplot(aes(x = year, y = deaths_by_year_cause/10000, color = cause)) +
    geom_line(size = 2) +
    gghighlight(cause == "Heart disease" |
                cause == "Cancer" |
                cause == "Alzheimer's disease") +
    scale_x_continuous(breaks = seq(2000, 2016, by = 2), expand = c(0, 0), limits = c(1999, 2017)) +
    scale_y_log10(expand = c(0, 0), breaks = c(3, 5, 10, 20, 30, 60, 100),
                  limits = c(2, 100)) + 
    # color-blind-friendly
    scale_color_OkabeIto() + 
    theme_classic(base_size = 15) +
    guides(color = "none") +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(face = "italic")) +
    labs(title = "Causes of Death Across Years",
         subtitle = "Heart disease and cancer are top 2 causes of deaths.\nAlzheimer's disease increases rapidly.",
         x = "Year",
         y = "No. of deaths (in 10k)",
         caption = "Source: Centers for Disease Control and Prevention ",
         color = "")

```

## Conclusion

This plot is intended to policty-makers and/or researchers to see the overall changes of causes of death across years. It seems heart disease, cancer and Alzheimer's disease are three causes that need most research resources in disease intervention and prevention, as first two disease are the top two distinct causes of death over years and Alzheimer's disease increases its amount rapidly. 




